---
title: "Station08 Annual Mean Temperature: Part I — Linear Trend and Residual Diagnostics"
subtitle: "Station 08 — Computational Methods in Econometrics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 3
    code_folding: show
    self_contained: true
editor_options:
  markdown:
    wrap: 72
---
# Purpose and how to use this document

This Part I document estimates a linear trend in the supplied annual temperature record and examines its residuals. It includes gap-aware plots, autocorrelation diagnostics, a Durbin–Watson Monte Carlo test, and a Breusch–Pagan test. Structural-change estimation, bootstrap inference, and the Monte Carlo comparison remain future work in Parts II and III.

This document contains the statistical code, explanations, tables, figures, interpretation and independent checks all in one place.


```{r setup, include=FALSE}
input_path <- knitr::current_input(dir = TRUE)
project_root <- normalizePath(dirname(input_path), mustWork = TRUE)
stopifnot(file.exists(file.path(project_root, "Station08.csv")))
artifact_root <- file.path(project_root, "work", "main_rmd_run")
dir.create(file.path(artifact_root, "knitr_figures"), recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = TRUE, error = FALSE, cache = FALSE,
  fig.align = "center", fig.path = paste0(file.path(artifact_root, "knitr_figures"), "/")
)
options(knitr.graphics.rel_path = FALSE)
```

# 1 Model Specification and Notes

This report starts with the model
$$y_i=\beta_0 + \beta_1t_i + \varepsilon_i, \qquad i=1, \ldots,n,$$
where $i$ is observation position and $t_i$ is the transformed calendar years:
$$t_i = (\mathrm{Year}_i - 1900)/10.$$
**On rescaling time:** The year-1900 centering is retained. The intercept is the fitted linear model temperature at 1900, and the slope is expressed in degrees Celsius per decade. Centering changes the intercept's reference year but not the slope or fitted curve. Appendix 1 explains this choice.

**On data completeness** There are 226 observations covering 261 calendar years from 1753-2013, with 35 missing years across five gaps. The largest gap is **1981- 2009**, comprising 29 unobserved years. Observation position and calendar time therefore differ. We retain every supplied observation, preserve calendar spacing,  and explain which diagnostics use each kind of lag.


# 0.2 Data Loading and Diagnosis of the Gap Structure

```{r load-data}
set.seed(20260914)

library(ggplot2)
library(lmtest)   # used only for post-hoc cross-checking against standard implementations; formal conclusions rely on our own implementation

dat <- read.csv(file.path(project_root, "Station08.csv"))

# Sort by year, keep only observations where both Year and Temperature are non-missing
dat <- dat[order(dat$Year), ]
dat <- dat[complete.cases(dat[, c("Year", "Temperature")]), ]
n <- nrow(dat)

cat("Number of observations n =", n, "\n")
cat("First year:", min(dat$Year), " Last year:", max(dat$Year), "\n")
```

```{r gap-detection}
# ---- Gap diagnostics: locate every "year jump" ----
full_years     <- seq(min(dat$Year), max(dat$Year))
missing_years  <- setdiff(full_years, dat$Year)
# year difference between consecutive observations
gap_len        <- diff(dat$Year)
# a difference > 1 indicates a gap
gap_start_idx  <- which(gap_len > 1)

gap_table <- data.frame(
  from      = dat$Year[gap_start_idx],
  to        = dat$Year[gap_start_idx + 1],
  n_missing = gap_len[gap_start_idx] - 1
)

print(gap_table)
cat("\nTotal missing years:", length(missing_years), "\n")
cat("Number of gaps:", nrow(gap_table), "\n")
big_gap <- gap_table[which.max(gap_table$n_missing), ]
cat("Largest gap:", big_gap$n_missing, "years, from", big_gap$from, "to", big_gap$to, "\n")
```

The data confirm **5 gaps totaling 35 missing years**, the vast majority
of which is the 29-year gap from 1981–2009. This means that two
observations that are adjacent in "sequence position" are not
necessarily adjacent in calendar time — this is the fundamental reason
why the ACF and DW test below require special handling.

To ensure that every subsequent time-connected plot is not misleading,
we construct a "contiguous year segment" label here: only observations
adjacent within the same segment will be connected by a line.

```{r segment-and-t}
# Increment the segment id by 1 whenever the year jumps by more than 1 year; used to avoid drawing lines across gaps
dat$segment <- c(1, cumsum(diff(dat$Year) > 1) + 1)

# Centered, decade-scaled time index
dat$t <- (dat$Year - 1900) / 10
```

# 1. Exploratory Analysis

```{r explore-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = Temperature)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_line(aes(group = segment), linewidth = 0.4, colour = "steelblue4") +
  geom_point(size = 0.8, colour = "steelblue4") +
  labs(
    title = "Geneva annual mean temperature (1753-2013)",
    subtitle = "Grey shading = largest data gap (1981-2009, no observations)",
    x = "Year", y = "Annual mean temperature (°C)"
  ) +
  theme_minimal(base_size = 12)
```

```{r explore-stats}
cat("Temperature range:", round(range(dat$Temperature), 2), "\n")
cat("5 coldest years:\n")
print(head(dat[order(dat$Temperature), c("Year", "Temperature")], 5))
cat("5 warmest years:\n")
print(tail(dat[order(dat$Temperature), c("Year", "Temperature")], 5))
cat("Standard deviation before 1900:", round(sd(dat$Temperature[dat$Year < 1900]), 3),
    "  Standard deviation from 1900 onward:", round(sd(dat$Temperature[dat$Year >= 1900]), 3), "\n")
```

**Main features:**

- **Direction of the trend**: The series shows an overall slow upward
  trend, rising from a mean level of about 9.3°C in the mid-18th century
  to levels above 10.5°C in 2010–2013, consistent with the broad
  direction of global warming.
- **Interannual variability**: Year-to-year fluctuations are clearly
  larger than the change in the trend itself (standard deviation of
  about 0.6–0.7°C); the temperature can swing by more than 1°C between
  two adjacent years, suggesting that a purely deterministic trend is
  unlikely to explain all of the variation.
- **Unusual observations**: The coldest years are concentrated in the
  19th century (1851, 1808, 1829, etc., with a minimum of 7.65°C); the
  warmest years, apart from 1834, 1947, and 1961, are mainly
  concentrated in 2011–2012 after the gap (reaching 11.73°C and 11.29°C
  respectively), forming two prominent high points at the tail of the
  series.
- **Visible changes in the behaviour of the series**: The most
  noticeable "change in behaviour" is actually **a structural absence of
  data** rather than a change in the temperature itself — there are no
  observations at all for the full 29 years from 1981–2009, which is
  precisely the period widely recognized as the fastest phase of global
  warming. Therefore, the apparent "jump" in the plot from 1980 directly
  to noticeably warmer values around 2010 could reflect genuine
  accelerated warming, or could simply be an artifact of the sample
  happening to pick up warmer years at both ends — the current data
  cannot distinguish between these two explanations. We return to this
  point in the overall assessment in Section 7.

# 2. OLS Estimation of the Linear Trend

```{r ols-fit}
trend_model <- lm(Temperature ~ t, data = dat)
summary(trend_model)

beta0_hat <- coef(trend_model)[1]
beta1_hat <- coef(trend_model)[2]

cat("Estimated intercept (fitted temperature around year 1900):", round(beta0_hat, 3), "°C\n")
cat("Estimated slope:", round(beta1_hat, 4), "°C per decade\n")
cat("Converted to per century:", round(beta1_hat * 10, 3), "°C per century\n")
```

```{r ols-plot, fig.width=8, fig.height=4.5}
dat$fitted_linear <- fitted(trend_model)

ggplot(dat, aes(x = Year, y = Temperature)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_line(aes(group = segment), linewidth = 0.4, colour = "grey40") +
  geom_point(size = 0.8, colour = "grey40") +
  geom_line(aes(y = fitted_linear), colour = "firebrick", linewidth = 1) +
  labs(title = "Geneva temperature series with OLS linear trend",
       x = "Year", y = "Annual mean temperature (°C)") +
  theme_minimal(base_size = 12)
```

**Interpretation:** $\hat\beta_0 = 9.617°C$ is the model-implied
"average" temperature level around the year 1900;
$\hat\beta_1 = 0.0337°C$ per decade (about 0.337°C per century)
represents the estimated long-run linear warming rate.
$R^2 \approx 0.105$, indicating that the deterministic trend explains
only about 10.5% of the total variation, with the vast majority of
interannual variability remaining in the residuals — consistent with the
impression from the exploratory analysis that "interannual variability
far exceeds the change due to the trend."

**On the impact of the gaps:** This step's point estimation is **not
affected by the gaps** — as long as the missingness is unrelated to the
temperature level itself (a reasonable assumption: missing data more
plausibly reflects a break in recording/digitization rather than data
being removed because the temperature was unusual), OLS remains unbiased
and consistent even with an unevenly-spaced observation sequence.
However, two limitations should be noted: (i) the effective sample size
drops to 226, so the standard errors are correspondingly larger; (ii)
since 1981–2009 is entirely blank, the estimated slope is essentially
determined by the contrast between the "1753–1980" and "2010–2013"
segments — if the true trend has accelerated or undergone a structural
change in recent decades, this single linear slope cannot detect it, and
can only provide an "overall average" warming rate. As required by the
assignment, we do not draw conclusions about the statistical
significance of the slope at this stage; we only report and interpret
the estimated results.

# 3. Residual Diagnostics

```{r residuals}
dat$residual    <- residuals(trend_model)
dat$residual_sq <- dat$residual^2
```

## Residuals over Time

```{r resid-time-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = residual)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(aes(group = segment), linewidth = 0.4) +
  geom_point(size = 0.8) +
  labs(title = "Residuals from the linear trend model", x = "Year", y = "Residual (°C)") +
  theme_minimal(base_size = 12)
```

The residuals fluctuate around 0 without any obvious systematic
divergence or convergence over time, but there is a tendency for runs of
same-signed residuals (a positive residual tends to be followed by
another positive residual), suggesting possible positive serial
correlation — we check this formally below using the ACF and the DW
test.

## Residual Autocorrelation Function (ACF) — Gap-Sensitive Step, Corrected

**Why the original `acf()` is problematic:** By default, `acf()` treats
the input vector as an **evenly spaced** time series, defining "lag k"
by the **sequence position** of the observations rather than the true
calendar year. In our data, the 178th observation is the year 1980 and
the 179th observation is the year 2010 — if `acf()` is applied directly
to the residual vector, it would treat these two residuals as a "lag 1"
pair, even though they are actually 30 calendar years apart. This
incorrectly mixes a long-distance relationship into the estimate of
"first-order autocorrelation," distorting the shape of the ACF.

**Correction:** Place the residuals back onto the complete 1753–2013
calendar grid, filling missing years with `NA`, construct a `ts` object,
and then compute the ACF with `na.action = na.pass` — this way R
computes autocorrelation based on the **true calendar lag**, correctly
skipping the "spurious lag 1" that crosses the gap rather than treating
it as a genuine first-order relationship.

```{r acf-compare, fig.width=9, fig.height=4}
# Naive approach (uncorrected, shown only for comparison to illustrate the problem)
acf_naive <- acf(dat$residual, plot = FALSE)

# Corrected approach: place back on the full calendar grid + na.action = na.pass
res_full <- rep(NA_real_, length(full_years))
res_full[match(dat$Year, full_years)] <- dat$residual
res_ts   <- ts(res_full, start = min(full_years), frequency = 1)
acf_gap  <- acf(res_ts, na.action = na.pass, lag.max = 20, plot = FALSE)

par(mfrow = c(1, 2))
plot(acf_naive, main = "Uncorrected ACF (ignores calendar gaps)")
plot(acf_gap,   main = "Gap-sensitive ACF (na.action = na.pass)")

cat("Uncorrected lag-1 autocorrelation:", round(acf_naive$acf[2], 3), "\n")
cat("Gap-sensitive lag-1 autocorrelation:", round(acf_gap$acf[2], 3), "\n")
```

The lag-1 autocorrelation coefficients from the two methods are fairly
close (about 0.29 versus 0.30), indicating that in this particular case
the gaps do not distort the shape of the ACF very severely (most gaps
are only 1–3 years, and there is only one "spurious pair" that truly
spans 29 years). But **in principle** this discrepancy could be
arbitrarily large, depending on the location and size of the gaps, so in
series with noticeable gaps it is prudent to insist on a
"calendar-sensitive" computation of the ACF. Both versions show positive
autocorrelation near lag 1 that gradually decays as the lag order
increases, which is a typical signature of short-run positive serial
correlation.

## Residual Variance over Time

```{r variance-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = residual_sq)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_point(size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  labs(
    title = "Squared residuals over time",
    subtitle = "Caution: the smooth line is not supported by any observations inside the shaded region",
    x = "Year", y = expression(hat(epsilon)[t]^2)
  ) +
  theme_minimal(base_size = 12)
```

The scatter itself uses the true Year as the x-axis and is unaffected by
unequal spacing; but the overlaid loess smooth, when passing through the
29-year gap from 1981–2009, is essentially **interpolation/extrapolation
with no data support whatsoever** — the noticeable upward tick at the
end of the curve is almost entirely driven by the 4 points from
2010–2013 alone, and should not be interpreted as "the variance was also
rising during the gap." When interpreting this plot, the smooth line
inside the shaded region should be treated only as a visual aid
connecting the two segments, not as evidence.

## Normal QQ Plot of Residuals

```{r qq-plot, fig.width=5, fig.height=5}
qqnorm(dat$residual, main = "Normal QQ plot of residuals")
qqline(dat$residual)
```

In the QQ plot, apart from a few points that fall slightly below the
reference line at the far left (the most negative theoretical
quantiles), the vast majority of points lie tightly along the diagonal,
broadly consistent with the normality assumption. The mild deviation in
the left tail corresponds to the handful of unusually cold 19th-century
years mentioned earlier (1851, 1808, 1805, etc.). This step **is not
affected by the gaps**, because the QQ plot only concerns the marginal
distribution of the residuals, which does not depend on the
observations' time order or spacing.

# 4. Testing for First-Order Residual Autocorrelation: Durbin–Watson Monte Carlo Test

**Why the original DW statistic also needs correction:** The original
code

``` r
dw_stat <- function(residuals) sum(diff(residuals)^2) / sum(residuals^2)
```

uses `diff()` to take term-by-term differences of the residual vector,
which has exactly the same problem as `acf()` — it treats **sequentially
adjacent** residual pairs as a "first-order" relationship, so the
difference between the 1980 and 2010 residuals is included in the
numerator, even though this is really a "30th-order" relationship and
should not appear in a first-order autocorrelation test.

**Correction:** We redefine
$$d = \frac{\sum_{(s,s+1)\in \mathcal{C}} (\hat\varepsilon_{s+1} - \hat\varepsilon_s)^2}{\sum_t \hat\varepsilon_t^2}$$
where $\mathcal{C}$ contains only those observation pairs that are
**truly adjacent on the calendar** (year difference exactly equal to 1).
The denominator still uses the sum of squares of all residuals (the
denominator does not involve the notion of "adjacency" and is
unaffected). This $d$ can be written as the quadratic form
$d = e'Ae / e'e$, where $A = D'D$ and $D$ is a matrix that takes
differences only over the valid adjacent pairs in $\mathcal{C}$ — this
follows exactly the same construction logic as the original DW
statistic, only restricting "which differences are taken" to truly
adjacent years.

```{r dw-define}
# Naive DW (uncorrected, shown only for comparison)
dw_naive_stat <- function(e) sum(diff(e)^2) / sum(e^2)
d_obs_naive <- dw_naive_stat(dat$residual)

# ---- Gap-sensitive DW: only compute differences over pairs that are truly calendar-adjacent (year difference = 1) ----
valid_pairs <- which(diff(dat$Year) == 1)      # indices of valid "adjacent pairs"
m_pairs <- length(valid_pairs)

Dmat <- matrix(0, nrow = m_pairs, ncol = n)
for (k in seq_len(m_pairs)) {
  i <- valid_pairs[k]
  Dmat[k, i]     <- -1
  Dmat[k, i + 1] <-  1
}
A_gap <- crossprod(Dmat)   # A = D'D

dw_gap_stat <- function(e) as.numeric((t(e) %*% A_gap %*% e) / (t(e) %*% e))
d_obs <- dw_gap_stat(dat$residual)

cat("Naive DW statistic (uncorrected):", round(d_obs_naive, 4), "\n")
cat("Gap-sensitive DW statistic:", round(d_obs, 4), "\n")
cat("Valid calendar-adjacent year pairs used:", m_pairs, "/", n - 1, "sequence-adjacent pairs\n")
```

The two values are quite close in this example (1.4147 versus 1.3865),
because only 5 of the 225 sequence-adjacent pairs are not genuine
calendar-adjacent pairs, so removing them correctly has a limited
impact; but this correction is necessary in principle, and the
discrepancy would be more pronounced with more numerous or larger gaps.
The Monte Carlo test below uniformly uses the **gap-sensitive version**
of $d$.

## Design of the Monte Carlo Test

Under the null hypothesis $H_0:$ no first-order autocorrelation, if
$\varepsilon \sim N(0,\sigma^2 I_n)$, then the residuals
$e = M\varepsilon$ ($M$ being the residual-maker matrix constructed from
the design matrix $X$). Since $d = e'Ae/e'e$ is a ratio of two quadratic
forms, $\sigma^2$ cancels out automatically, so the **exact
finite-sample distribution** of $d$ under $H_0$ depends only on $X$
(through $M$) and $A$ (through the gap structure), not on the unknown
$\beta,\sigma^2$ — this is precisely the theoretical basis that allows
us to simulate the exact null distribution using standard-normal
pseudo-errors (discussed in detail in Section 6). The key point here is:
the $A$ matrix used in the simulation must be the same $A$ used for
$d_{obs}$ (i.e., also constructed based on true calendar adjacency),
otherwise the simulated reference distribution and the observed
statistic would not correspond to the same statistic, invalidating the
test.

```{r dw-mc-test}
X <- model.matrix(trend_model)
H <- X %*% solve(t(X) %*% X) %*% t(X)
M <- diag(n) - H

set.seed(20260914)
B <- 9999
d_sim <- numeric(B)

for (b in 1:B) {
  eps_star <- rnorm(n)                     # standard-normal pseudo-errors under H0
  e_star   <- as.numeric(M %*% eps_star)   # corresponding OLS residuals
  d_sim[b] <- dw_gap_stat(e_star)          # using the exact same A matrix as for d_obs
}

alpha <- 0.05
crit_lower <- quantile(d_sim, alpha / 2)
crit_upper <- quantile(d_sim, 1 - alpha / 2)

p_lower_adj <- (sum(d_sim <= d_obs) + 1) / (B + 1)
p_upper_adj <- (sum(d_sim >= d_obs) + 1) / (B + 1)
p_val_dw <- 2 * min(p_lower_adj, p_upper_adj, 0.5)

cat("--- Monte Carlo Durbin-Watson Test (gap-sensitive version) ---\n")
cat("d_obs:", round(d_obs, 4), "\n")
cat("MC lower critical value (2.5%):", round(crit_lower, 4), "\n")
cat("MC upper critical value (97.5%):", round(crit_upper, 4), "\n")
cat("MC two-sided p-value:", round(p_val_dw, 4), "\n")

# Cross-check: lmtest::dwtest uses the naive (uncorrected) statistic + asymptotic normal approximation, for reference only
dwtest(trend_model)
```

**Results and conclusion:** $d_{obs} = 1.3865$. Because the
finite-sample distribution of $d$ under the null need not be symmetric
around 2, we obtain separate lower and upper critical values of $1.7031$
and $2.2241$ (rather than simply computing the upper one as $4$ minus
the lower one). The **rejection region** is $d < 1.7031$ or
$d > 2.2241$. The observed value $1.3865$ falls below the lower critical
value, and the MC two-sided p-value is $0.0002$; at the $\alpha=0.05$
level we reject the null hypothesis of "no first-order autocorrelation,"
providing strong evidence that the residuals exhibit **positive
first-order serial correlation**. This is fully consistent with the
graphical diagnostics (the runs of same-signed residuals in the residual
plot, and the positive value of the ACF at lag 1).

# 5. Testing for Heteroskedasticity: Breusch–Pagan Test

The auxiliary regression
$\hat\varepsilon_t^2 = \gamma_0 + \gamma_1 t + v_t$ tests
$H_0: \gamma_1 = 0$ (homoskedasticity) against $H_1: \gamma_1 \ne 0$.

**On the impact of the gaps:** This step is essentially **unaffected**
by the gaps — the auxiliary regression uses each observation's **true**
$t$ value as the regressor, and the formulas for $R^2$ and $BP=nR^2$
themselves do not rely on an equal-spacing assumption. The only indirect
effect of the gaps is that the effective sample size $n=226$ is smaller,
which affects the reliability of the asymptotic $\chi^2(1)$
approximation — but this is purely a sample-size issue, not the
structural "sequence-adjacent ≠ calendar-adjacent" problem that affects
DW/ACF (this is exactly the essential distinction discussed in Section
6).

```{r bp-test}
aux_model <- lm(residual_sq ~ t, data = dat)
R2_aux   <- summary(aux_model)$r.squared
BP_stat  <- n * R2_aux
p_val_bp <- 1 - pchisq(BP_stat, df = 1)

cat("Breusch-Pagan statistic BP:", round(BP_stat, 4), "\n")
cat("Degrees of freedom:", 1, "\n")
cat("BP p-value:", round(p_val_bp, 4), "\n")

# Cross-check
bptest(trend_model, varformula = ~ t, data = dat)
```

**Result:** $BP = 0.4173$, $p = 0.5183$, far greater than 0.05, so we
**cannot reject** the null hypothesis of homoskedasticity. The graphical
diagnostic (the variance-over-time plot in Section 3.3, setting aside
the unsupported extrapolation inside the shaded region) also shows no
systematic variance pattern over time — the two conclusions agree.

# 6. Comparing the Inferential Principles

Why is it valid to construct exact Monte Carlo critical values for the
**DW** statistic using the procedure "generate standard-normal
pseudo-errors → project through the $M$ matrix → compute the statistic,"
but the same idea cannot simply be transferred to the **BP** statistic?

- **The DW statistic is a "pivotal quantity":** $d = e'Ae/e'e$, where
  $e = M\varepsilon$ is a **linear** transformation of $\varepsilon$,
  and both $A$ and $M$ are fixed matrices that depend only on the design
  matrix $X$ (i.e., on the known year/gap structure), independent of the
  unknown parameters $\beta,\sigma^2$. As long as $\varepsilon$ is
  assumed normal, the **exact finite-sample distribution** of $d$ under
  $H_0$ depends only on $X$, and can be computed exactly (or
  approximated exactly by simulation) for any $n$ or sample design —
  this is precisely what makes the "MC test" valid in the Dufour sense:
  the simulated reference distribution *is* the true null distribution,
  not an approximation.

- **The BP statistic is fundamentally an asymptotic (large-sample)
  test:** $BP = nR^2_{aux}$ is a Lagrange-multiplier-type statistic
  whose auxiliary regression uses the **squared** residuals
  ($\hat\varepsilon_t^2$) as the dependent variable — a **nonlinear**
  (quadratic) transformation of $\varepsilon$. Even under the assumption
  that $\varepsilon$ is normal, $\hat\varepsilon_t^2$ is no longer
  normally distributed (it is approximately chi-square-type and skewed);
  the classical "$\chi^2(1)$" reference distribution is obtained via a
  large-sample central limit theorem approximation, which holds strictly
  only as $n\to\infty$. Its **finite-sample** exact distribution
  generally does not have a closed form depending only on $X$ as neatly
  as DW's does, and additionally depends on higher moments of the error
  distribution (such as kurtosis) that are unknown. In other words, the
  whole appeal of the BP test lies in not imposing strong assumptions on
  the specific error distribution (it is a relatively "robust"
  large-sample test); if we were to force a normality assumption purely
  in order to simulate an exact distribution, we would be giving up BP's
  main advantage — and even then, there is no simple guarantee, as there
  is for DW, that the simulated distribution equals the true
  finite-sample null distribution.

In short: **The exact MC test for DW is valid because it is a ratio of
quadratic forms that is linear in and scale-invariant with respect to**
$\varepsilon$, whose null distribution is "pivotal"; BP involves a
nonlinear transformation of $\varepsilon$, and its classical reference
distribution is itself an asymptotic approximation rather than an exact
result — it lacks the same pivotal property, so the same simulation
recipe cannot simply be carried over.

# 7. Overall Assessment of the Initial Model

Bringing the graphical and formal diagnostics together:

1.  **Clear direction, modest magnitude of trend**: OLS estimates a
    linear warming trend of about 0.034°C per decade (equivalently
    0.337°C per century), consistent with the exploratory analysis's
    impression of an "overall slow upward" pattern; $R^2\approx 0.105$,
    indicating that the deterministic trend explains only a small
    fraction of the total variation.
2.  **Normality broadly holds**: The QQ plot shows the residuals are
    approximately normally distributed, with only a mild deviation at
    the most extreme cold years, which is not a serious concern.
3.  **Homoskedasticity broadly holds**: The BP test ($p=0.518$) and the
    variance-over-time plot both find no evidence of a systematic change
    in variance over time (the apparent upward tick inside the shaded
    region is unsupported by data and should be disregarded).
4.  **The independence assumption is clearly violated**: The ACF and the
    Monte Carlo DW test ($d=1.3865$, $p=0.0002$) consistently indicate
    significant positive first-order autocorrelation in the residuals.
    This means the standard errors, t-values, and p-values reported by
    default in `lm()` (based on the classical formulas assuming i.i.d.
    errors) **are not trustworthy at this stage** — I guess this is also
    exactly why the assignment asks us not to draw conclusions about the
    significance of the trend coefficient yet: before the residual
    autocorrelation has been addressed, any significance inference based
    on classical OLS standard errors could be biased (typically
    understating the true standard error, thereby inflating apparent
    significance).
5.  **The data gaps themselves are a non-negligible limitation**: The
    largest gap happens to cover 1981–2009 (the period of fastest modern
    warming), meaning the slope estimate in this model is essentially
    determined by the contrast between the "pre-gap" segment and the
    "short post-gap tail," rather than being supported by a genuinely
    continuous, year-by-year observed curve. The sample data cannot
    distinguish between "the entire two-hundred-plus-year period does
    approximately follow a linear warming trend" and "warming has
    accelerated / undergone a structural change in recent decades."

**Conclusion:** The simple linear trend model provides a useful but
**rough** preliminary description — it correctly captures the direction
of long-term warming, and the normality and homoskedasticity of the
residuals are broadly defensible, but (i) the residuals exhibit
significant serial correlation, violating the independence assumption
required for classical OLS inference, and (ii) the structural gaps in
the data, particularly the nearly 30-year gap, further undermine our
confidence in the assumed "linear" functional form. Together, these
findings suggest that the linear trend model can only serve as a
**starting point** for subsequent analysis, rather than as a final
adequate description.

# Appendix 1: Choosing the Centric Year

This is a compromise between Interpretability and statistical
rationality. First of all, 1900 is a whole year that all readers are
familiar with and requires no explanation, making it convenient for
communication and memorization. Secondly, it is not too far from the
true sample mean (1866) and falls within a relatively dense interval of
the data distribution; therefore, although the standard error of the
intercept is not the smallest, it is still significantly better than
when using end-point years, and the estimate remains quite stable.
Moreover, in the context of climate change, the “early 20th century”
itself serves as a practical point of reference (often used as a
pre-industrial or early industrial benchmark), making it easier to align
with readers’ background knowledge than a mean year calculated purely
mathematically.

# Appendix 2: Impact of Data Gaps on This Analysis — Summary

| Component | Affected? | Approach adopted in this report |
|----|----|----|
| Exploratory plot | Visual level | Line drawn only within contiguous year segments + shading marks the largest gap |
| OLS point estimates $\hat\beta_0,\hat\beta_1$ | Essentially unaffected | No modification needed; interpretation notes the limited ability to identify "recent trend" |
| Residuals-over-time plot | Visual level | Same segment-based treatment as the exploratory plot |
| Residual ACF | **Substantive** | Missing years filled with NA to construct a `ts` object, using `na.action = na.pass` |
| Variance-over-time plot (loess) | Visual level | Gap shaded, with a text note that the smooth line is unsupported by data inside the gap |
| QQ plot | Unaffected | No modification needed |
| DW statistic and MC test | **Substantive** | $d$ redefined to use only differences between calendar-truly-adjacent residual pairs; the MC simulation matrix $A$ modified accordingly |
| BP test | Essentially unaffected | No modification needed |
| Overall model assessment | Conclusion level | Explicitly notes that the large gap weakens our ability to identify "whether a recent trend change exists" |
