---
title: "Station08 Annual Mean Temperature: Linear Trends, Structural Change and Bootstrap Inference"
subtitle: "Station 08 — Computational Methods in Econometrics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 3
    code_folding: show
    self_contained: true
params:
  recompute_simulation: false
  workers: 4
editor_options: 
  markdown: 
    wrap: 72

---
# Purpose and how to use this document

This Part I document estimates a linear trend in the supplied annual temperature record and examines its residuals. It includes gap-aware plots, autocorrelation diagnostics, a Durbin–Watson Monte Carlo test, and a Breusch–Pagan test. Structural-change estimation, bootstrap inference, and the Monte Carlo comparison remain future work in Parts II and III.

This document contains the statistical code, explanations, tables, figures, interpretation and independent checks all in one place.


```{r setup, include=FALSE}
input_path <- knitr::current_input(dir = TRUE)
project_root <- normalizePath(dirname(input_path), mustWork = TRUE)
stopifnot(file.exists(file.path(project_root, "Station08.csv")))
artifact_root <- file.path(project_root, "work", "main_rmd_run")
dir.create(file.path(artifact_root, "knitr_figures"), recursive = TRUE, showWarnings = FALSE)
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = TRUE, error = FALSE, cache = FALSE,
  fig.align = "center", fig.path = paste0(file.path(artifact_root, "knitr_figures"), "/")
)
options(knitr.graphics.rel_path = FALSE)
fmt <- function(value, digits = 4) {
  formatC(value, digits = digits, format = "f")
}
```

# 1 Model Specification and Notes

This report starts with the model
$$y_i=\beta_0 + \beta_1t_i + \varepsilon_i, \qquad i=1, \ldots,n,$$
where $i$ is observation position and $t_i$ is the transformed calendar years:
$$t_i = (\mathrm{Year}_i - 1900)/10.$$
**On rescaling time:** The year-1900 centering is retained. The intercept is the fitted linear model temperature at 1900, and the slope is expressed in degrees Celsius per decade. Centering changes the intercept's reference year but not the slope or fitted curve. Appendix 1 explains this choice.

**On data completeness** There are 226 observations covering 261 calendar years from 1753-2013, with 35 missing years across five gaps. The largest gap is **1981- 2009**, comprising 29 unobserved years. Observation position and calendar time therefore differ. We retain every supplied observation, preserve calendar spacing,  and explain which diagnostics use each kind of lag.


# 2 Data Loading and Diagnosis of the Gap Structure

```{r load-data}
master_seed <- 20260926L
set.seed(master_seed)

library(ggplot2)

dat <- read.csv(file.path(project_root, "Station08.csv"))
stopifnot(identical(names(dat), c("Year", "Temperature")),
          nrow(dat) == 226L, !anyNA(dat),
          all(is.finite(dat$Year)), all(is.finite(dat$Temperature)),
          !anyDuplicated(dat$Year))

# Sort by year, validate supplied data instead of silently dropping observations.
dat <- dat[order(dat$Year), ]
stopifnot(all(diff(dat$Year)>0))
n <- nrow(dat)

cat("Number of observations n =", n, "\n")
cat("First year:", min(dat$Year), " Last year:", max(dat$Year), "\n")
```

```{r gap-detection}
# ---- Gap diagnostics: locate every "year jump" ----
full_years     <- seq(min(dat$Year), max(dat$Year))
missing_years  <- setdiff(full_years, dat$Year)
# year difference between consecutive observations
gap_len        <- diff(dat$Year)
# a difference > 1 indicates a gap
gap_start_idx  <- which(gap_len > 1)

gap_table <- data.frame(
  from      = dat$Year[gap_start_idx],
  to        = dat$Year[gap_start_idx + 1],
  n_missing = gap_len[gap_start_idx] - 1
)

print(gap_table)
cat("\nTotal missing years:", length(missing_years), "\n")
cat("Number of gaps:", nrow(gap_table), "\n")
big_gap <- gap_table[which.max(gap_table$n_missing), ]
cat("Largest gap:", big_gap$n_missing, "years, from", big_gap$from, "to", big_gap$to, "\n")
```

The data confirm **5 gaps totaling 35 missing years**, the vast majority
of which is the 29-year gap from 1981–2009. This means that two
observations that are adjacent in "sequence position" are not
necessarily adjacent in calendar time — this is the fundamental reason
why the ACF and DW test below require special handling.

To ensure that every subsequent time-connected plot is not misleading,
we construct a "contiguous year segment" label here: only observations
adjacent within the same segment will be connected by a line.

```{r segment-and-t}
# Increment the segment id by 1 whenever the year jumps by more than 1 year; used to avoid drawing lines across gaps
dat$segment <- c(1, cumsum(diff(dat$Year) > 1) + 1)

# Centered, decade-scaled time index
dat$t <- (dat$Year - 1900) / 10
```

# 3. Exploratory Analysis

```{r explore-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = Temperature)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_line(aes(group = segment), linewidth = 0.4, colour = "steelblue4") +
  geom_point(size = 0.8, colour = "steelblue4") +
  labs(
    title = "Station08 annual mean temperature (1753-2013)",
    subtitle = "Grey shading = largest data gap (1981-2009, no observations)",
    x = "Year", y = "Annual mean temperature (°C)"
  ) +
  theme_minimal(base_size = 12)
```

```{r explore-stats}
cat("Temperature range:", round(range(dat$Temperature), 2), "\n")
cat("5 coldest years:\n")
print(head(dat[order(dat$Temperature), c("Year", "Temperature")], 5))
cat("5 warmest years:\n")
print(tail(dat[order(dat$Temperature), c("Year", "Temperature")], 5))
cat("Standard deviation before 1900:", round(sd(dat$Temperature[dat$Year < 1900]), 3),
    "  Standard deviation from 1900 onward:", round(sd(dat$Temperature[dat$Year >= 1900]), 3), "\n")
```

**Main features:**

- **Direction of the fitted trend:** Temperatures show considerable annual variability and a modest positive full-sample linear slope. This station alone cannot establish a global trend or its cause.
- **Interannual variability:** The sample standard deviation is about 0.70°C. Year-to-year variation is substantial relative to the fitted long-run change.
- **Unusual observations:** The coldest observed year is 1851, at about 7.65°C; the warmest is 2011, at 11.73°C. These observations are retained. Part II includes a descriptive check omitting the coldest year.
- **Missing decades:** The four 2010–2013 observations follow the 29-year absence of data. They do not show whether warming was gradual, accelerated, or interrupted during the gap. We do not infer a temperature discontinuity from the absence of observations.


# 4. OLS Estimation of the Linear Trend

```{r ols-fit}
trend_model <- lm(Temperature ~ t, data = dat)
summary(trend_model)

print(coef(trend_model))
cat("R-squared:", round(summary(trend_model)$r.squared, 4), "\n")

beta0_hat <- coef(trend_model)[1]
beta1_hat <- coef(trend_model)[2]

cat("Estimated intercept (fitted temperature around year 1900):", round(beta0_hat, 3), "°C\n")
cat("Estimated slope:", round(beta1_hat, 4), "°C per decade\n")
cat("Converted to per century:", round(beta1_hat * 10, 3), "°C per century\n")

```

```{r ols-plot, fig.width=8, fig.height=4.5}
dat$fitted_linear <- fitted(trend_model)

ggplot(dat, aes(x = Year, y = Temperature)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_line(aes(group = segment), linewidth = 0.4, colour = "grey40") +
  geom_point(size = 0.8, colour = "grey40") +
  geom_line(aes(y = fitted_linear), colour = "firebrick", linewidth = 1) +
  labs(title = "Geneva temperature series with OLS linear trend",
       x = "Year", y = "Annual mean temperature (°C)") +
  theme_minimal(base_size = 12)
```

**Interpretation:** $\hat\beta_0 = 9.617°C$ is the model-implied
"average" temperature level around the year 1900;
$\hat\beta_1 = 0.0337°C$ per decade (about 0.337°C per century)
represents the estimated long-run linear warming rate.
$R^2 \approx 0.105$, indicating that the deterministic trend explains
only about 10.5% of the total variation, with the vast majority of
interannual variability remaining in the residuals — consistent with the
impression from the exploratory analysis that "interannual variability
far exceeds the change due to the trend."

**On the impact of the gaps:** Using actual calendar year preserves the meaning of the slope; OLS can be computed for an unevenly spaced fixed design. Missing years nevertheless affect which periods are represented and can affect estimates, precision and interpretation. Exogeneity of the errors and the missingness mechanism cannot be verified from the CSV. The 222 observations through 1980 also inform the fitted slope; it is not determined solely by a contrast between two endpoint groups. A single linear slope averages across the observed record and does not identify what happened within the missing decades. As required in Part I, we do not draw a slope-significance conclusion here.


# 3. Residual Diagnostics

```{r residuals}
dat$residual    <- residuals(trend_model)
dat$residual_sq <- dat$residual^2
```

## Residuals over Time

```{r resid-time-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = residual)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(aes(group = segment), linewidth = 0.4) +
  geom_point(size = 0.8) +
  labs(title = "Residuals from the linear trend model", x = "Year", y = "Residual (°C)") +
  theme_minimal(base_size = 12)
```

The residuals fluctuate around zero, with runs of similarly signed values and low-frequency departures from the linear fit. These patterns suggest residual dependence and possible inadequacy of the constant-trend mean. We inspect both observation-lag and calendar-lag ACFs, then use Monte Carlo calibration for the assigned DW statistic.

## Residual Autocorrelation Function (ACF) — Observation and Calendar Lags



**Why the lag definition matters:** Applied to the observed residual vector, `acf()` measures lag by row position. It is a valid description of dependence between observed rows, but a one-observation lag is not always a one-calendar-year lag. For example, observations `r match(1980, dat$Year)` and `r match(2010, dat$Year)` correspond to 1980 and 2010, thirty calendar years apart.

**Calendar-lag comparison:** Place residuals on the full calendar grid, insert `NA` for unobserved years, and use `na.action = na.pass`. This uses available pairs for each calendar lag without imputing temperatures. With missing values, the resulting estimates need not define a valid autocorrelation sequence; the usual plot bounds are descriptive rather than calibrated inference. Both ACFs are retained, and their lag meanings are stated explicitly.


```{r acf-compare, fig.width=9, fig.height=4}
# Naive approach (uncorrected, shown only for comparison to illustrate the problem)
acf_naive <- acf(dat$residual, lag.max = 20, plot = FALSE)

# Corrected approach: place back on the full calendar grid + na.action = na.pass
res_full <- rep(NA_real_, length(full_years))
res_full[match(dat$Year, full_years)] <- dat$residual
res_ts   <- ts(res_full, start = min(full_years), frequency = 1)
acf_gap  <- acf(res_ts, na.action = na.pass, lag.max = 20, plot = FALSE)

par(mfrow = c(1, 2))
plot(acf_naive, main = "ACF by observation lag")
plot(acf_gap,   main = "ACF by calendar-year lag")

cat("Observation-lag 1 autocorrelation:", round(acf_naive$acf[2], 3), "\n")
cat("Calendar-lag 1 autocorrelation:", round(acf_gap$acf[2], 3), "\n")
par(mfrow = c(1, 1))

```

The first observation-lag and calendar-lag autocorrelations are `r fmt(as.numeric(acf_naive$acf[2]), 3)` and `r fmt(as.numeric(acf_gap$acf[2]), 3)`. They are close here because most observed pairs are consecutive calendar years. Both indicate positive short-range residual dependence. Their similarity in this dataset does not make the two lag definitions interchangeable in general.


## Residual Variance over Time

```{r variance-plot, fig.width=8, fig.height=4.5}
ggplot(dat, aes(x = Year, y = residual_sq)) +
  annotate("rect", xmin = big_gap$from, xmax = big_gap$to,
           ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.6) +
  geom_point(size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  labs(
    title = "Squared residuals over time",
    subtitle = "Caution: the smooth line is not supported by any observations inside the shaded region",
    x = "Year", y = expression(hat(epsilon)[t]^2)
  ) +
  theme_minimal(base_size = 12)
```

The scatter uses actual calendar years. The loess curve is a descriptive smooth, and its portion inside the long gap is not supported by observations from those years. The final four observations can influence its endpoint behavior; the plot cannot establish how variance evolved during the gap. No formal conclusion about constant variance follows from this smooth alone.


## Normal QQ Plot of Residuals

```{r qq-plot, fig.width=5, fig.height=5}
qqnorm(dat$residual, main = "Normal QQ plot of residuals")
qqline(dat$residual)
```

The QQ plot is broadly linear in its center, with some tail departures. This is compatible with a Gaussian approximation but does not verify normality. Time spacing does not enter the QQ construction, although the plotted marginal distribution still reflects only the observed years. Gaussian errors remain an explicit reference assumption for the DW Monte Carlo calibration.

# 4. Testing for First-Order Residual Autocorrelation: Durbin–Watson Monte Carlo Test
**Primary assigned statistic:** The handout defines

$$d_{\mathrm{obs}}=\frac{\sum_{i=2}^n(\widehat e_i-\widehat e_{i-1})^2}{\sum_{i=1}^n\widehat e_i^2}.$$

Its adjacency is by observation position. We implement this definition as the primary test and interpret it accordingly. Calendar gaps affect its meaning, but restricting its numerator to calendar-adjacent pairs would define a different statistic rather than implement the assigned one.

**Retained calendar-gap sensitivity:**  original variant is kept:

$$d_{\mathrm{calendar}}=\frac{\sum_{i:\ s_{i+1}-s_i=1}(\widehat e_{i+1}-\widehat e_i)^2}{\sum_i\widehat e_i^2}.$$

It includes only genuinely adjacent calendar years. Both statistics can be written as $e'Ae/e'e$, with different difference matrices. Each is calibrated using its own formula and the same observed regression design.


```{r dw-define}
# Naive DW (uncorrected, shown only for comparison)
dw_naive_stat <- function(e) sum(diff(e)^2) / sum(e^2)
d_obs_naive <- dw_naive_stat(dat$residual)

# ---- Gap-sensitive DW: only compute differences over pairs that are truly calendar-adjacent (year difference = 1) ----
valid_pairs <- which(diff(dat$Year) == 1)      # indices of valid "adjacent pairs"
m_pairs <- length(valid_pairs)

Dmat <- matrix(0, nrow = m_pairs, ncol = n)
for (k in seq_len(m_pairs)) {
  i <- valid_pairs[k]
  Dmat[k, i]     <- -1
  Dmat[k, i + 1] <-  1
}
A_gap <- crossprod(Dmat)   # A = D'D

dw_gap_stat <- function(e) as.numeric((t(e) %*% A_gap %*% e) / (t(e) %*% e))
d_obs_gap <- dw_gap_stat(dat$residual)
d_obs <- d_obs_naive  # primary statistic: assignment definition

cat("Assigned observation-adjacency DW:", round(d_obs_naive, 4), "\n")
cat("Gap-sensitive DW statistic:", round(d_obs_gap, 4), "\n")
cat("Valid calendar-adjacent year pairs used:", m_pairs, "/", n - 1, "sequence-adjacent pairs\n")

```

The assigned statistic is `r fmt(d_obs)`, and the calendar-adjacent variant is `r fmt(d_obs_gap)`. There are `r m_pairs` calendar-adjacent pairs among the `r n-1` observation-adjacent pairs. Both are reported, with the assigned statistic governing the main Part I conclusion.

## Design of the Monte Carlo Test

Under the specified null $\varepsilon\sim N(0,\sigma^2I_n)$, the fitted residuals are $e=M_X\varepsilon$, where $M_X=I-X(X'X)^{-1}X'$. Regression coefficients disappear after projection and error scale cancels in $e'Ae/e'e$. Thus the null law is determined by the fixed design and the chosen difference matrix.

We retain projection and simulation loop. Each standard-normal draw is evaluated using **both** DW definitions. Separate lower and upper empirical critical values give the two tails; they are not obtained by reflecting one tail around two. The previous Rmd’s interpolated type-7 critical-value quantiles are retained. Simulation approximates the specified finite-sample Gaussian law, rather than an unrestricted null of every process without first-order correlation.


```{r dw-mc-test}
X <- model.matrix(trend_model)
H <- X %*% solve(t(X) %*% X) %*% t(X)
M <- diag(n) - H

set.seed(master_seed)
B <- 9999L
d_sim <- d_sim_gap <- numeric(B)

for (b in seq_len(B)) {
  eps_star <- rnorm(n)
  e_star <- as.numeric(M %*% eps_star)
  d_sim[b] <- dw_naive_stat(e_star)   # assigned observation-adjacency statistic
  d_sim_gap[b] <- dw_gap_stat(e_star) # retained calendar-adjacent sensitivity
}

alpha <- 0.05
crit_lower <- unname(quantile(d_sim, alpha / 2, type = 7))
crit_upper <- unname(quantile(d_sim, 1 - alpha / 2, type = 7))
p_lower_adj <- (sum(d_sim <= d_obs) + 1) / (B + 1)
p_upper_adj <- (sum(d_sim >= d_obs) + 1) / (B + 1)
p_val_dw <- min(1, 2 * min(p_lower_adj, p_upper_adj))

crit_gap <- unname(quantile(d_sim_gap, c(alpha / 2, 1 - alpha / 2), type = 7))
p_val_dw_gap <- min(1, 2 * min(
  (sum(d_sim_gap <= d_obs_gap) + 1) / (B + 1),
  (sum(d_sim_gap >= d_obs_gap) + 1) / (B + 1)
))
dw_results <- data.frame(
  Definition = c("Assigned: observation adjacency", "Sensitivity: calendar adjacency"),
  Statistic = c(d_obs, d_obs_gap),
  Lower_2_5_percent = c(crit_lower, crit_gap[1]),
  Upper_97_5_percent = c(crit_upper, crit_gap[2]),
  Equal_tail_p_value = c(p_val_dw, p_val_dw_gap)
)
knitr::kable(dw_results, digits = 4)

# Optional packaged cross-check uses its own statistic/reference and is not the primary calibration.
if (requireNamespace("lmtest", quietly = TRUE)) {
  print(lmtest::dwtest(trend_model))
}

```

**Results and conclusion:** The assigned statistic is `r fmt(d_obs)`, with lower and upper cutoffs `r fmt(crit_lower)` and `r fmt(crit_upper)` and equal-tail p-value `r fmt(p_val_dw)`. It falls below the lower cutoff. The calendar-adjacent sensitivity also rejects in the positive-dependence direction, with p-value `r fmt(p_val_dw_gap)`. These results agree with the residual and ACF plots. Rejection challenges the specified independent Gaussian linear-model errors; departures in the mean specification can also contribute to residual patterns.


# 5. Testing for Heteroskedasticity: Breusch–Pagan Test

The auxiliary regression

$$\widehat e_i^2=\gamma_0+\gamma_1t_i+v_i$$

tests for a **linear variance trend**, using the requested statistic $BP=nR^2$ and nominal $\chi^2_1$ reference. A zero time coefficient is a narrow diagnostic null; non-rejection does not establish homoskedasticity against every possible variance pattern.

**On gaps and calibration:** The regression uses actual transformed calendar year and does not require equally spaced rows for its algebra. Missing periods still limit interpretation, and the detected residual dependence weakens the usual asymptotic chi-square reference. This limitation is separate from the choice of observation versus calendar adjacency in DW and ACF.


```{r bp-test}
aux_model <- lm(residual_sq ~ t, data = dat)
R2_aux   <- summary(aux_model)$r.squared
BP_stat  <- n * R2_aux
p_val_bp <- pchisq(BP_stat, df = 1, lower.tail = FALSE)

cat("Breusch-Pagan statistic BP:", round(BP_stat, 4), "\n")
cat("Degrees of freedom:", 1, "\n")
cat("BP p-value:", round(p_val_bp, 4), "\n")

# Previous cross-check: bptest(trend_model, varformula = ~ t, data = dat)
# Optional cross-check; no dependency on lmtest for the main result.
if (requireNamespace("lmtest", quietly = TRUE)) {
  print(lmtest::bptest(trend_model, varformula = ~ t, data = dat))
}

```

**Result:** $BP=$ `r fmt(BP_stat)` and the nominal p-value is `r fmt(p_val_bp)`. We do not reject a linear variance trend at 5%. This is compatible with the descriptive plot but does not prove constant variance, especially given the residual dependence.


# 6. Comparing the Inferential Principles

The main distinction is the **reference assumptions**, not whether residuals are squared.

- **DW under a specified Gaussian null:** Residual projection removes the regression coefficients, and the ratio of quadratic forms removes scale. Conditional on the observed design and difference matrix, standard-normal simulation therefore generates the specified finite-sample reference law. A finite simulation estimates its quantiles; separate corrected tail counts give the reported Monte Carlo p-value.
- **BP with the requested chi-square reference:** The usual $nR^2$ calibration is asymptotic and needs its regularity conditions. Under a broader homoskedastic null, the error distribution, higher moments and dependence are not fully specified.
- **Gaussian Monte Carlo BP is also possible:** If the same independent Gaussian error law is imposed, project standard-normal draws, square their residuals and recompute the auxiliary $R^2$. Multiplying all residuals by an error scale multiplies all squared residuals by its square and leaves this $R^2$ unchanged. Thus Gaussian simulation can also calibrate BP conditional on the design. The nonlinear squaring step does not prevent this. The issue is that imposing Gaussian independence is stronger than assuming homoskedasticity alone.

Here we retain the assignment's asymptotic chi-square BP calculation and qualify it in light of dependence rather than treating non-rejection as proof of constant variance.


# 7. Overall Assessment of the Initial Model
Bringing the graphical and formal diagnostics together:

1. **Positive descriptive slope:** The full-sample linear estimate is `r fmt(beta1_hat)`°C per decade, with $R^2=$ `r fmt(summary(trend_model)$r.squared, 3)`. This is a modest average fitted change, not a Part I significance claim.
2. **Gaussian reference remains an assumption:** The QQ plot is broadly linear in its center with tail departures; it does not verify normality.
3. **No detected linear variance trend:** The nominal BP p-value is `r fmt(p_val_bp)`. Non-rejection does not establish homoskedasticity, and dependence qualifies the reference law.
4. **Residual dependence matters:** Both DW variants and both ACFs indicate positive residual dependence. Ordinary independent-error OLS standard errors are not a reliable basis for the requested later inference without addressing this issue. A richer mean may reduce, but need not eliminate, the residual pattern.
5. **Missing periods constrain interpretation:** Actual calendar spacing is retained, but there is no direct evidence about temperatures during 1981–2009. The missingness mechanism and station measurement history are unknown.

**Conclusion for Part I:** The linear model is a useful starting description. Its residual patterns motivate a continuous change in slope and a bootstrap that can retain dependence. Parts II–III below carry out that inference and evaluate its finite-sample behavior.

# 8. Structural Change and Boostrap Inderence (Part II)

```{r part2-setup}
year <- dat$Year
y <- dat$Temperature
x <- dat$t
m0 <- trend_model
e0 <- dat$residual
seed <- master_seed
B_test <- 9999L
B_ci <- 4999L
B_dw <- B
simulation_M <- 1000L
simulation_B <- 499L
dw <- d_obs
dwcript<-c(crit_lower, crit_upper)
dwp<-p_val_dw
bp<- BP_stat
bpp <- p_val_bp
ncores<- parallel::detectCores()
if (is.na(ncores)) ncores <- 1L
stopifnot(is.logical(params$recompute_simulation),
          length(params$recompute_simulation) == 1L,
          !is.na(params$recompute_simulation),
          is.numeric(params$workers), length(params$workers) == 1L,
          is.finite(params$workers), params$workers >= 1)
workers <- min(as.integer(params$workers), ncores)
